The Way Forward in Parallelizing Dynamic Languages

### 2014-04-29 talk/icooolps2014/position-paper.tex
add some text to the introduction

\title{Title Text}
\subtitle{Position Paper, ICOOOLPS'14}

@@

\section{Introduction}

\subsection*{Issue}
efficiently supporting multi-CPU usage on dynamic languages that were designed with GIL semantics in
mind

(supporting (large) atomic blocks for synchronization)

\subsection*{Our Position}
Current solutions for replacing the GIL include STM, HTM, and
fine-grained locking. STM is usually too slow, HTM very limited, and
locking suffers from complexity that makes it hard to implement
correctly. We argue that the best way forward is still STM and that
its performance problem can be solved.

@@

\section{Discussion}
\paragraph{dynamic language VM problems}

>>>>

\title{The Way Forward in Parallelizing Dynamic Languages}
\subtitle{Position Paper, ICOOOLPS'14}

@@

\section{Introduction}
In a world where computers get more and more cores and single-thread
performance increases less and less every year, many dynamic languages
have a problem. While there is certainly a lot of popularity around
languages like Python and Ruby, their ability to make use of multiple
cores is somewhat limited. For ease of implementation they chose to
use a single, global interpreter lock (GIL) to synchronize the
execution of code in multiple threads. While this is a
straight-forward way to eliminate synchronization issues in the
interpreter, it prevents parallel execution. Code executed in multiple
threads will be serialized over this GIL so only one thread executes
at a time.

There exist several solutions and work-arounds to remove or avoid the
GIL in order to benefit from multiple cores. We are going to discuss
several of them and try to find the best way forward. The first
approach uses fine-grained locking to replace the single GIL. Then
there are shared-nothing models that use for example multiple
processes with multiple interpreters and explicit message
passing. Finally, one can also directly replace the GIL with
transactional memory (TM), either software-based (STM) or
hardware-based (HTM).

The approach that wins in the end should perform similarly for
single-threaded execution as compared to the GIL and be able to
execute code in parallel on multiple cores. Furthermore, we will also
take into account the compatibility to existing code that already uses
threads for concurrency, as well as the changes that are required to
the interpreter itself.

These requirements are not easy to meet. We argue that STM is the
overall winner. While it has a big performance problem currently, it
gets more points in all the other categories. We think that it is the
only solution that also enables a better synchronization mechanism to
the application in the form of atomic blocks.

%% \subsection{Issue}
%% The issue that we want to discuss is how to efficiently support
%% multi-core parallel execution of code in dynamic languages that were
%% designed with GIL semantics in mind.

%% Furthermore, a solution to this problem should also bring better
%% synchronization mechanism with it...

%% (supporting (large) atomic blocks for synchronization)

%% \subsection{Our Position}
%% Current solutions for replacing the GIL include STM, HTM, and
%% fine-grained locking. STM is usually too slow, HTM very limited, and
%% locking suffers from complexity that makes it hard to implement
%% correctly. We argue that the best way forward is still STM and that
%% its performance problem can be solved.

@@

\section{Discussion}
(why do we have the GIL? it makes the interpreter thread-safe)
(now there is code relying on the semantics (atomicity of bytecode instructions))
(synchronization with locks for application-level synchronization is still needed. So GIL is useless for that)

\paragraph{dynamic language VM problems}
